\documentclass[11pt]{article}
\usepackage[hyperref]{acl}
\usepackage{times}
\usepackage{latexsym}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{inconsolata}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}

\title{Comparing Chunking Strategies for Retrieval-Augmented Question Answering}

\author{
  Assylkhan Geniyat \\
  Boston University \\
  \texttt{kuromiqo@bu.edu} \And
  Pablo Bello \\
  Boston University \\
  \texttt{pbello@bu.edu} \And
  Jaime Bernal \\
  Boston University \\
  \texttt{jaimebm@bu.edu} \AND
  Batyrkhan Baimukhanov \\
  Boston University \\
  \texttt{batyr@bu.edu}
}

\date{}

\begin{document}
\maketitle

\begin{abstract}
This empirical project studies how document chunking affects retrieval-augmented question answering. Our proposal asked whether chunking strategy changes end-to-end QA accuracy, and whether the best strategy depends on query complexity. To investigate this, we built a controlled RAG pipeline that holds the embedding model and dense retriever fixed while varying only the chunking function. The current comparison includes fixed-size token chunking, recursive chunking, sentence-based chunking, and semantic chunking, where the semantic splitter acts as a dynamic content-aware chunker rather than a fixed window. We now run the generator with \texttt{mistralai/Mistral-7B-Instruct-v0.3} served through vLLM on SCC. Experiments on SQuAD v2 pseudo-documents and HotpotQA distractor examples still show a clear chunking effect. On SQuAD, semantic chunking gives the best token-level F1, while on HotpotQA recursive chunking is strongest. A no-retrieval baseline remains weak overall, confirming that the task is retrieval-sensitive rather than purely parametric. Absolute answer scores are lower than in our earlier local prototype, which suggests that generator interface and prompt design interact strongly with chunking quality. We therefore treat the current rerun as both a validation of the chunking trend and a useful signal for the final-stage prompt and model analysis.
\end{abstract}

\section{Introduction}

Retrieval-augmented generation (RAG) answers questions by retrieving external evidence and passing it to a language model \citep{lewis2020rag}. In a standard pipeline, however, retrieval is performed over chunks rather than full documents. Chunking is therefore not a minor preprocessing choice: it determines what evidence stays together, how likely the retriever is to recover it, and whether the generator receives coherent support or fragmented context.

This project follows the empirical plan in our proposal. The central question is: \textit{which chunking strategy maximizes end-to-end QA accuracy in a fixed RAG pipeline, and does the answer differ between relatively local questions and more complex multi-hop questions?}

Formally, let a corpus be $D$, let a question be $q$, and let a chunking function be $c : D \rightarrow \{d_1, \dots, d_n\}$. After chunking, each chunk is embedded and indexed. At inference time, the system retrieves the top-$K$ chunks for $q$, concatenates them into a prompt, and produces an answer $\hat{a}$. The optimization target is end-to-end QA quality rather than retrieval in isolation:
\[
c^* = \arg\max_c \; \mathbb{E}_{(q,a)} \left[ F_1(\hat{a}, a) \right].
\]

This problem is interesting for two reasons. First, chunking is widely treated as an implementation detail, even though recent studies suggest that it can substantially affect retrieval behavior. Second, retrieval quality and answer quality do not always move together: a system may retrieve locally relevant text yet still fail if the evidence is incomplete, noisy, or badly fragmented. These observations motivated our proposal to evaluate chunking with both end-to-end QA metrics and retrieval diagnostics.

Simple baselines are informative here. A no-retrieval system should fail whenever the answer depends on the supplied corpus rather than parametric memory. Likewise, naive fixed windows can fail in opposite ways: large chunks may include distractors, while very small chunks may separate evidence that should remain together. Our current experiments already reveal both effects.

\section{Related Work}

RAG typically combines neural retrieval with text generation by retrieving passages and conditioning a generator on them \citep{lewis2020rag}. Dense Passage Retrieval \citep{karpukhin2020dpr} established dense retrieval as a strong backbone for open-domain question answering, which makes it a natural fixed retriever for our chunking study.

For evaluation, we use the same dataset pair proposed originally. SQuAD~2.0 \citep{rajpurkar2018squad2} provides mostly local answer extraction over Wikipedia passages, while HotpotQA \citep{yang2018hotpotqa} stresses multi-hop reasoning and evidence composition. This pairing lets us test whether chunking behaves differently for local factoid questions and multi-document reasoning.

Our chunking motivation also comes from recent engineering and benchmarking analyses. \citet{smith2024chunking} show that retrieval quality can vary substantially across chunking policies. NVIDIA's engineering study argues that chunking should be judged end-to-end rather than only through retrieval recall \citep{nvidia2024chunking}. \citet{qu2024semanticchunking} further question whether semantic chunking always justifies its extra computational cost. Our contribution is narrower but controlled: we isolate chunking as the only experimental variable inside a reproducible QA pipeline.

\section{Methods}

The implemented system follows the proposal closely. For each chunking strategy, the pipeline is:
\begin{enumerate}
  \item Chunk each document.
  \item Embed every chunk with the same encoder.
  \item Build the same FAISS dense index.
  \item Retrieve the top 4 chunks for each question.
  \item Pass the retrieved context to the same generator.
\end{enumerate}

The chunking function is the only variable.

\subsection{Chunking Strategies}

We currently implement the four chunking families from the proposal:
\begin{itemize}
  \item \textbf{Fixed token chunking:} \texttt{fixed\_128}, \texttt{fixed\_256}, and \texttt{fixed\_512}, with overlaps of 19, 38, and 76 tokens respectively.
  \item \textbf{Recursive chunking:} \texttt{recursive\_256}, implemented with LangChain's recursive splitter.
  \item \textbf{Sentence chunking:} \texttt{sentence\_256}, implemented with spaCy sentence segmentation and greedy packing to a token budget.
  \item \textbf{Semantic chunking:} \texttt{semantic\_256}, implemented by embedding consecutive sentences and splitting when cosine similarity falls below a threshold. In the reported run, the similarity threshold is 0.72; in the newer codebase, we also support a minimum-length guard to stabilize very short chunks. In practice, this is our dynamic chunker: chunk length varies with local semantic continuity rather than a fixed token budget.
\end{itemize}

We also implemented auxiliary \texttt{chonkie\_recursive\_256} and \texttt{chonkie\_semantic\_256} variants to check whether the same high-level chunking idea behaves similarly across implementations.

\subsection{Models and Software}

For the current midway rerun, the rest of the pipeline is fixed as follows:
\begin{itemize}
  \item Embedding model: \texttt{sentence-transformers/all-MiniLM-L6-v2}
  \item Retriever: FAISS inner-product search over normalized dense embeddings
  \item Generator: \texttt{mistralai/Mistral-7B-Instruct-v0.3} served through vLLM on SCC
  \item Retrieval depth: top 4 chunks
  \item Generation limits: 768 input tokens and 32 new tokens
  \item Embedding batch size: 32
  \item Runtime: local retrieval/chunking with remote OpenAI-compatible generation on SCC
\end{itemize}

This is now aligned with the original proposal rather than a deviation from it. The broader codebase also supports BM25, hybrid retrieval, reranking, supporting-document coverage, and bootstrap confidence intervals, but the midway analysis keeps the ablation focused on chunking. All chunking hyperparameters are fixed in JSON config files and reused across runs.

\subsection{Baseline}

The proposal required at least one simple baseline. Our current baseline is \textbf{parametric-only QA}, in which the generator answers the question without retrieved context. This tests whether retrieval is actually necessary and gives a clear lower bound for retrieval-based systems.

\section{Experiments}

\subsection{Datasets}

We use the two datasets proposed originally:
\begin{itemize}
  \item \textbf{SQuAD v2:} 60 answerable validation questions are sampled from a shuffled 500-example candidate pool. Paragraphs with the same title are concatenated into article-level pseudo-documents to make chunking non-trivial. The resulting corpus has 35 pseudo-documents.
  \item \textbf{HotpotQA distractor:} 30 validation questions are used, with the induced document collection built from the distractor and supporting titles in each example. The resulting corpus has 291 documents.
\end{itemize}

The current subsets are small, but they are already large enough to expose clear differences between chunking strategies.

\subsection{Evaluation}

Our evaluation follows the proposal:
\begin{itemize}
  \item \textbf{End-to-end QA:} Exact Match (EM) and token-level F1
  \item \textbf{Retrieval:} Recall@4 and Precision@4
  \item \textbf{Efficiency:} average chunk size, number of indexed chunks, and latency-related statistics
\end{itemize}

For SQuAD, Recall@4 reduces to single-document recall. For HotpotQA, it is the fraction of gold supporting documents recovered in the top 4 retrieved chunks, which is more informative than an ``any-hit'' definition for multi-hop questions.

\section{Results}

\input{generated/midway_mistral/midway_tables.tex}

The midway results still support the proposal's main hypothesis that chunking matters. The parametric-only baseline is weak overall, so the task remains retrieval-sensitive rather than purely solvable from parametric memory alone.

On SQuAD, \texttt{semantic\_256} is the strongest retrieval-based setting at 24.4 F1 and ties for the best Recall@4 at 95.0. \texttt{fixed\_128} has the best retrieval precision at 77.1, but lower answer quality at 21.0 F1. This pattern still suggests that local factoid questions benefit from semantically coherent boundaries, not just from retrieving a locally pure top-4 set. Exact match is noisier in this rerun: the parametric baseline reaches 5.0 EM despite much lower F1, which indicates that exact string matching is unstable under the chat-style Mistral prompting interface and should not be read in isolation.

HotpotQA is more diagnostic. \texttt{recursive\_256} is strongest at 18.8 F1, while doc-level Recall@4 tops out at 76.7 rather than saturating at 100.0. This gives a more realistic picture of multi-hop retrieval quality. Semantic chunking no longer ties for the top spot here, which suggests that once the generator changes, boundary quality alone is not enough: success also depends on how well the prompt format lets the model compose evidence across documents. The stronger-model rerun therefore validates the chunking trend only partially. Semantic chunking remains best on the more local dataset, while recursive chunking is more robust on the multi-hop dataset.

\section{Discussion and Project Plan}

The current failure profile matches the taxonomy laid out in the proposal. Some errors are clear retrieval failures, especially when no answer-bearing chunk appears in the top 4. Others are context-fragmentation failures, where evidence is partially retrieved but split in an unhelpful way. A third category is now visible in the SCC rerun: generator-interface failures, where retrieval looks adequate but the instruction-tuned chat model still answers vaguely or misses the exact span. On SQuAD, the strongest semantic results suggest that preserving local coherence is more important than maximizing top-4 purity alone. On HotpotQA, recursive chunking is more robust, which implies that evidence composition and prompt grounding are now the main bottlenecks once retrieval reaches roughly three quarters of the supporting documents.

The remaining work is concrete:
\begin{enumerate}
  \item \textbf{Scale the experiments.} Increase the evaluation subsets so the comparison is statistically more stable.
  \item \textbf{Tune the stronger SCC generator.} Now that \texttt{Mistral-7B-Instruct-v0.3} is running on SCC, tune the QA prompt format and compare it against the earlier local prototype.
  \item \textbf{Strengthen retrieval diagnostics.} Add supporting-document coverage and a cleaner separation between retrieval failure and generation failure, especially for HotpotQA.
  \item \textbf{Deepen error analysis.} Add more qualitative examples showing when chunking helps and when it fragments evidence.
  \item \textbf{Integrate the broader retrieval stack.} Use the already implemented BM25, hybrid, and reranked settings to test whether chunking effects persist under stronger retrieval.
\end{enumerate}

Relative to the proposal, the project has stayed on the same conceptual trajectory. The main new insight is that switching to a stronger 7B-class model does not automatically improve end-to-end QA under the same retrieval setup; chunking quality, prompt design, and generator interface interact more strongly than our initial local prototype suggested.

\bibliography{references}
\bibliographystyle{acl_natbib}

\end{document}
